---
title: "KM_Survival"
output: html_document
date: "2023-06-21"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries.
```{r}
library(tidyverse)
library(survival)
library(lubridate)
library(ggthemes)
library(readxl)
library(writexl)
library(janitor)
library(survminer)
```

Importing data and formatting.
```{r}
#Importing data
fate_clean_km_2021_22 <- read_xlsx("/Volumes/Samsung_T5/BOBWHITE_DATA/Clean/fate_clean_2021_22.xlsx") %>%
  mutate(fate_date = ymd(fate_date),
         date_added = ymd(date_added)) 

#Taking a quick look. Event = 1 is a mortality, 0 is right censored.
fate_clean_km_2021_22 %>%
  view()


```

Beginning with staggered entry KM.
```{r}
#First, we will specify when our survival year begins. We began adding birds the week of 20 October 2021.
start_date <- as.Date("2021-10-1")

#Next, we will create a column stating when each bird was added to the study. Birds added Oct 20 2021 will be day 0, birds added in Spring (3/14) will be day 145
fate_clean_km_2021_22$day_added <- as.numeric(fate_clean_km_2021_22$date_added - start_date)

#Appears to have worked.
fate_clean_km_2021_22 %>%
  view()

#Next, we will create a column "days_alive" stating how many days each bird was alive since 20 October 2021 (start time).
fate_clean_km_2021_22$days_alive <- as.numeric(fate_clean_km_2021_22$fate_date - start_date)

#Appears to have worked.
fate_clean_km_2021_22 %>%
  view()

```


Fitting our model. We'll add vertical lines indicating winter and annual survival.
```{r}
#fitting survival curve.
km_surv_fit <- survfit(Surv(time = fate_clean_km_2021_22$day_added, time2 = fate_clean_km_2021_22$days_alive, event = fate_clean_km_2021_22$fate) ~1)

#Looking at summary, this just seems to dived birds into two intervals -- birds added in Fall and birds added in Spring. It still 110 birds at time = 0, as well.
summary(km_surv_fit)

#Plotting our survival curve. This isn't what we want!
plot(km_surv_fit,
     main = "KM Survival curve",
     xlab = "Number of days after start date (1 October 2021)",
     ylab = "Survival")
abline(v = 181, col = "midnight blue", lwd = 2, lty = 2)
abline(v = 365, col = "red", lwd = 2, lty = 2)
```

Now, let's create a data frame so we can extract our survival % at March 31st 2022 (winter survival), and Sept 30 2022 (Annual survival)
```{r}
# Creating the dataframe with confidence interval
KM_DF_2021_22 <- data.frame(time = km_surv_fit$time,
                            surv = km_surv_fit$surv,
                            lower = km_surv_fit$lower,
                            upper = km_surv_fit$upper)

#Getting winter survival. We have to use which.min() here, because survival % is only calculated when an event occurs. This will get us to the closest possible day of an event at March 31st.
winter_surv_2021_22 <- KM_DF_2021_22$surv[which.min(abs(KM_DF_2021_22$time - 181))]
lower_winter_2021_22 <- KM_DF_2021_22$lower[which.min(abs(KM_DF_2021_22$time - 181))]
upper_winter_2021_22 <- KM_DF_2021_22$upper[which.min(abs(KM_DF_2021_22$time - 181))]
print(c(winter_surv_2021_22, lower_winter_2021_22, upper_winter_2021_22)) #53% seems reasonable.

#Getting annual survival. Same situation as above with which.min().
annual_surv_2021_22 <- KM_DF_2021_22$surv[which.min(abs(KM_DF_2021_22$time - 365))]
lower_annual_2021_22 <- KM_DF_2021_22$lower[which.min(abs(KM_DF_2021_22$time - 365))]
upper_annual_2021_22 <- KM_DF_2021_22$upper[which.min(abs(KM_DF_2021_22$time - 365))]
print(c(annual_surv_2021_22, lower_annual_2021_22, upper_annual_2021_22)) #23% seems reasonable, a little higher than we had on excel spreadsheet.
```

Ok, now lets see if we can do the same thing, but for Summer survival, which begins on April 1st 2022.
```{r}
#First, let's filter out birds that died before then.
fate_km_summer_2022 <- fate_clean_km_2021_22 <- read_xlsx("/Volumes/Samsung_T5/BOBWHITE_DATA/Clean/fate_clean_2021_22.xlsx") %>%
  mutate(fate_date = ymd(fate_date),
         date_added = ymd(date_added)) %>%
  filter(fate_date > ymd("2022-3-30"))

#Looks like that worked.
fate_km_summer_2022 %>%
  view()

#Since all our birds for summer survival were "added" on 1 April 2022, lets create a column stating that.
fate_km_summer_2022 <- fate_km_summer_2022 %>%
  mutate(day_added_summer = as.numeric(0))

#Now, we will specify our Summer start date as April 1 2022.
start_date_summer <- ymd("2022-4-1")


#Next, we will create a column "days_alive_summer" stating how many days each bird was alive after 1 April 2022
fate_km_summer_2022$days_alive_summer <- as.numeric(fate_km_summer_2022$fate_date - start_date_summer)

#Appears to have worked.
fate_km_summer_2022 %>%
  view()
```

Fitting our model for summer survival.
```{r}
#fitting survival curve.
km_surv_fit_summer_2022 <- survfit(Surv(time = fate_km_summer_2022$day_added_summer, time2 = fate_km_summer_2022$days_alive_summer, event = fate_clean_km_2021_22$fate) ~1)

#Looking at summary, this just seems to dived birds into two intervals -- birds added in Fall and birds added in Spring. It still 110 birds at time = 0, as well.
summary(km_surv_fit_summer_2022)

#Plotting our survival curve. This isn't what we want!
plot(km_surv_fit_summer_2022,
     main = "KM Summer Survival curve",
     xlab = "Number of days after start date (1 April 2022)",
     ylab = "Survival")
abline(v = 182, col = "red", lwd = 2, lty = 2)

```

Now, let's create our dataframe for summer survival.
```{r}
# Creating the dataframe
KM_DF_Summer_2022 <- data.frame(time = km_surv_fit_summer_2022$time,
                      surv = km_surv_fit_summer_2022$surv)

#Getting winter survival. We have to use which.min() here, because rows only exist when an event occurs. This will get us to the closest possible day of an event at March 31st.
Summer_survival_2022 <- KM_DF_Summer_2022$surv[which.min(abs(KM_DF_Summer_2022$time - 182))]
print(Summer_survival_2022) 
```

