---
title: "KM_Survival"
output: html_document
date: "2023-06-21"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries.
```{r}
library(tidyverse)
library(survival)
library(lubridate)
library(ggthemes)
library(readxl)
library(writexl)
library(janitor)
```

Importing data and formatting.
```{r}
#Importing data
fate_clean_km_2021_22 <- read_xlsx("/Volumes/Samsung_T5/BOBWHITE_DATA/Clean/fate_clean_2021_22.xlsx") %>%
  mutate(fate_date = ymd(fate_date),
         date_added = ymd(date_added)) 

#Taking a quick look. Event = 1 is a mortality, event = 0 is right censored.
fate_clean_km_2021_22 %>%
  view()
```

Beginning with staggered entry KM.
```{r}
#First, we will specify when our survival year begins (1 Oct 2021). We began adding birds the week of 20 October 2021...maybe that should be the start date, instead?
start_date <- ymd("2021-10-1")

#Next, we will create a numeric column stating when each bird was added to the study. Birds added Oct 20 2021 will be day 19, birds added in Spring (14 March) will be day 164. 
fate_clean_km_2021_22$day_added <- as.numeric(fate_clean_km_2021_22$date_added - start_date)

#Appears to have worked.
fate_clean_km_2021_22 %>%
  view()

#Next, we will create a numeric column "days_alive" stating how many days each bird was alive since 1 October 2021 (start time).
fate_clean_km_2021_22$days_alive <- as.numeric(fate_clean_km_2021_22$fate_date - start_date)

#Appears to have worked.
fate_clean_km_2021_22 %>%
  view()

#Irrelevant to what we're doing here, but curious to see the mean days alive.
mean(fate_clean_km_2021_22$days_alive)
```


Fitting and plotting our model. We'll add vertical lines indicating winter and annual survival end dates.
```{r}
#fitting survival curve.
km_surv_fit <- survfit(Surv(time = fate_clean_km_2021_22$day_added, time2 = fate_clean_km_2021_22$days_alive, event = fate_clean_km_2021_22$fate) ~1)

#Looking at summary. Appears sound.
summary(km_surv_fit)

#Plotting our survival curve. Note that it is 100% between 1 Oct - 20 Oct, since we did not have any birds online.
plot(km_surv_fit,
     main = "KM Survival curve",
     xlab = "Number of days after start date (1 October 2021)",
     ylab = "Survival",
     xlim = c(0, 365))
#Below we add vertical lines at winter and annual survival end dates.
abline(v = 181, col = "midnight blue", lwd = 2, lty = 3)
abline(v = 365, col = "red", lwd = 2, lty = 2)
```

Now, let's create a data frame so we can extract our survival % at March 31st 2022 (winter survival), and Sept 30 2022 (Annual survival)
```{r}
# Creating the dataframe with confidence interval
KM_DF_2021_22 <- data.frame(time = km_surv_fit$time,
                            surv = km_surv_fit$surv,
                            lower = km_surv_fit$lower,
                            upper = km_surv_fit$upper)

#Getting winter survival. We have to use which.min() here, because new rows are only created when an event occurs. This will get us to the closest possible day of an event at March 31st.
winter_surv_2021_22 <- KM_DF_2021_22$surv[which.min(abs(KM_DF_2021_22$time - 181))]
lower_winter_2021_22 <- KM_DF_2021_22$lower[which.min(abs(KM_DF_2021_22$time - 181))]
upper_winter_2021_22 <- KM_DF_2021_22$upper[which.min(abs(KM_DF_2021_22$time - 181))]

#Below, we print, lower CI, winter survival, upper CI
print(c(lower_winter_2021_22, winter_surv_2021_22, upper_winter_2021_22)) #53% seems reasonable.

#Getting annual survival. Same situation as above with which.min().
annual_surv_2021_22 <- KM_DF_2021_22$surv[which.min(abs(KM_DF_2021_22$time - 365))]
lower_annual_2021_22 <- KM_DF_2021_22$lower[which.min(abs(KM_DF_2021_22$time - 365))]
upper_annual_2021_22 <- KM_DF_2021_22$upper[which.min(abs(KM_DF_2021_22$time - 365))]
print(c(lower_annual_2021_22, annual_surv_2021_22, upper_annual_2021_22)) #23% seems reasonable, a little higher than we had on excel spreadsheet.
```


Ok, now lets see if we can do the same thing, but for Summer survival, which begins on April 1st 2022.
```{r}
#First, let's filter out birds that died before then.
fate_km_summer_2022 <- fate_clean_km_2021_22 <- read_xlsx("/Volumes/Samsung_T5/BOBWHITE_DATA/Clean/fate_clean_2021_22.xlsx") %>%
  mutate(fate_date = ymd(fate_date),
         date_added = ymd(date_added)) %>%
  filter(fate_date > ymd("2022-3-30"))

#Looks like that worked.
fate_km_summer_2022 %>%
  view()

#Now, we will specify our Summer start date as April 1 2022.
start_date_summer <- ymd("2022-4-1")

#Since all our birds for summer survival were "added" on 1 April 2022, lets create a numeric column stating that.
fate_km_summer_2022 <- fate_km_summer_2022 %>%
  mutate(day_added_summer = as.numeric(0))


#Next, we will create a column "days_alive_summer" stating how many days each bird was alive after 1 April 2022.
fate_km_summer_2022$days_alive_summer <- as.numeric(fate_km_summer_2022$fate_date - start_date_summer)

#Appears to have worked.
fate_km_summer_2022 %>%
  view()
```

Fitting our model for summer survival.
```{r}
#fitting survival curve.
km_surv_fit_summer_2022 <- survfit(Surv(time = fate_km_summer_2022$day_added_summer, time2 = fate_km_summer_2022$days_alive_summer, event = fate_clean_km_2021_22$fate) ~1)

#Looking at summary. Appears sound.
summary(km_surv_fit_summer_2022)

#Plotting our survival curve. 
plot(km_surv_fit_summer_2022,
     main = "KM Summer Survival curve",
     xlab = "Number of days after start date (1 April 2022)",
     ylab = "Survival",
      xlim = c(0, 182))
#Adding vertical line at Sept 30 2022.
abline(v = 182, col = "red", lwd = 2, lty = 2)

```

Now, let's create our dataframe for summer survival.
```{r}
# Creating the dataframe with confidence interval
KM_DF_Summer_2022 <- data.frame(time = km_surv_fit_summer_2022$time,
                            surv = km_surv_fit_summer_2022$surv,
                            lower = km_surv_fit_summer_2022$lower,
                            upper = km_surv_fit_summer_2022$upper)

#Getting winter survival. We have to use which.min() here, because new rows are only created when an event occurs. This will get us to the closest possible day of an event at Sept 30th.
summer_survival_2022 <- KM_DF_Summer_2022$surv[which.min(abs(KM_DF_Summer_2022$time - 182))]
lower_summer_2022 <- KM_DF_Summer_2022$lower[which.min(abs(KM_DF_Summer_2022$time - 182))]
upper_summer_2022 <- KM_DF_Summer_2022$upper[which.min(abs(KM_DF_Summer_2022$time - 182))]

#Below, we print, lower CI, summer survival, upper CI
print(c(lower_summer_2022, summer_survival_2022, upper_summer_2022)) #45% seems a bit lower than we had on KM excel spreadsheet. Need to double check.
```

